Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-06-17 ref_response.tex

Despite using these relations to correct SN Ia peak brightnesses to within a dispersion of $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitors of SNe Ia is yet to be confirmed. While it is likely that SNe Ia are caused by mass transfer onto a WD from, or violent merger with, a companion star, multiple possible scenarios exist for the nature of that companion star (see \citealt{Maoz2014} for a review). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, or more exotic models such as the core-degenerate (CD) scenario which invoke WDs merging with the cores of massive stars \citep{Soker2019}.

@@

An indirect method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD describes the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review of the theory, and \citealt{Maoz2017} for a review of observations). The SD scenario produces a broad range of functional forms for the DTD, most of which fail to account for long delay times \citep{Graur2014}. On the other hand, most variants of the DD scenario predict a power-law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}.

@@

Instead of comparing \textit{volumetric} rates to the \textit{cosmic} SFH, it is also possible to estimate the SFH of individual galaxies through the modelling of their stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012}, \citet{Graur2013}, and \citet{Graur2015} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$.

@@

The majority of observational evidence based on studies of SN Ia populations thus points towards a DD scenario for most, if not all, SNe Ia. However, finding self-consistent progenitor and explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs (linked strongly with the SD scenario but also with many DD scenarios) find difficulty in reproducing the light curves of "normal" SNe Ia as well as "peculiar" objects (\citealt{Ropke2007,Sim2013,Blondin2017}; see \citealt{Maoz2014,Jha2019} for overviews). In recent years, attention directed towards explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations (primarily related to a DD scenario) has led to promising results \citep[e.g.][]{Shen2017,Shen2018,Townsley2019,Gronow2020,Shen2021} although they still struggle to match observations at late times in the light curve evolution \citep{Shen2021,Gronow2021}. An additional factor in support of the sub-$M_{\mathrm{Ch}}$ model is that the SN luminosity is related to the mass of the primary WD, which itself is likely to be related to its age (although this relation is probably complicated by other factors such as accretion rate, metallicity, and the composition of the companion), thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Rose2019,Nicolas2020}. With many models showing promising similarities to observations but each subject to its own drawbacks, it is becoming accepted that more than one progenitor scenario may contribute significantly to the overall population of "normal" SNe Ia; detailed observations are thus required in order to place constraints on the relative fractions of each possible progenitor channel.

In this work, we combine parts of the traditional methods and derive a new measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012}. We use the stellar mass assembly model of \citetalias{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

@@

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we use Bayesian inference. To sample the posterior we use the enhanced no-U-turn Sampler (NUTS) algorithm \citep{Betancourt2017}, which is a variant of Hamiltonian Monte-Carlo (HMC), implemented in the Stan programming language \citep{Carpenter2017}. We describe the fitting procedure in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.63\pm0.02$, which is consistent with the value found in weakly star-forming galaxies in the Supernova Legacy Survey by \citet{Sullivan2006}, but also in passive galaxies in SDSS by \citet{Smith2012}, and in all galaxies in the Lick Observatory Supernova Search (LOSS) by \citet{Li2011a}.

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit simultaneously by a single two-component DTD, with fixed slope $\beta = -1$ and a prompt fraction that is 1 at short delay times, 0 at large delay times, and a linear function of time between $t_1 =0.65$~Gyr (where tardy SNe switch on) and $t_2=0.85$~Gyr (where prompt SNe turn off).%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citetalias{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Graur2017,Rigault2018,Rose2019,Nicolas2020,Rose2021}. The corresponding DTD fit results in a steep power-law $\beta =-1.70^{+0.19}_{-0.10}$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. The difference in $\beta$ is significant at a level of $3.8\sigma$. This result provides an intriguing confirmation of that observed in 101 SDSS SNe Ia by \citet{Brandt2010} who found that low-$x_1$ SNe displayed much longer delay times than those with high $x_1$.

From the differing DTDs that describe SNe with low and high stretch values we infer that there are either multiple channels through which SN Ia explode which occur on differing timescales, or that the explosion mechanism evolves with progenitor age. Two scenarios, that are not necessarily mutually exclusive, that could lead to different DTD decay rates are as follows. 
The first such scenario is to assume that all SNe Ia come from the same initial population of stars, and thus that the progenitors of both low and high stretch SNe form at the same time. It is then necessary to invoke models in which the WD progenitors of low-stretch SNe evolve on longer timescales than those of high-stretch SNe, for example due to a different initial binary separation distribution or accretion rate. In this case, the DTD slopes would indeed take on the different values measured here.

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are mildly consistent with one another, with $t_1 =0.69^{+0.07}_{-0.13}$~Gyr and $t_2 = 0.81^{+0.13}_{-0.10}$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are reminiscent of the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while low-stretch SNe occurred with delay times $\gtrsim 2.4$~Gyr. The transition found in our analysis appears to occur at much shorter delay times than the several Gyrs found by \citet{Rose2019}. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but suggest that such a scenario is unlikely.

@@

One possibility is that sub-luminous SNe are more numerous in predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at delay times as long as $>6$~Gyr \citep{Panther2019}, and are typically sub-luminous with low stretch \citep{Gonzalez-Gaitan2014}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, their presence in high-mass hosts could explain the apparent lack of normal SNe Ia in the sample. We check this hypothesis briefly by re-examining the objects left out of our sample due to \texttt{SALT2} cuts. We find that 75\% of objects left out lie in hosts with $M_*>10^{10}$ M$_{\odot}$, while this percentage is 65\% for the objects that pass the cuts. However, of the objects in high-mass galaxies that fail the cuts, only 11 (10\%) fail due to their $x_1$ lying below $-3$, and a total of 20 have $x_1 <-2$ which is the regime populated by SN 1991bg-like objects which is well short of the factor of two by which the DTD model over-predicts the data in the highest mass bin. Another issue is that SN 1991bg-like SNe are generally fainter than normal SNe Ia and the massive galaxies in which they explode are typically bright, leading to a lower detection efficiency than that modelled in Section \ref{subsec:incompleteness_SNe}. A thorough investigation into the presence of SN Ia sub-types in the DES-SN data set is deferred to a future investigation.

>>>>

Despite using these relations to correct SN Ia peak brightnesses to within a dispersion of $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitors of SNe Ia is yet to be confirmed. While it is likely that SNe Ia are caused by mass transfer onto a WD from, or violent merger with, a companion star, multiple possible scenarios exist for the nature of that companion star (see \citealt{Maoz2014,Ruiter2020} for reviews). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, or more exotic models such as the core-degenerate (CD) scenario which invoke WDs merging with the cores of massive stars \citep{Kashi2011,Ilkov2012}.

@@

An indirect method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD describes the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review of the theory, and \citealt{Maoz2017} for a review of observations). The SD scenario produces a broad range of functional forms for the DTD, most of which fail to account for long delay times \citep{Graur2014}. On the other hand, most variants of the DD scenario predict a power-law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}, although see e.g.  \citet{Yungelson2017} for DD models that deviate from a $-1$ power-law slope.

@@

Instead of comparing \textit{volumetric} rates to the \textit{cosmic} SFH, it is also possible to estimate the SFH of individual galaxies through the modelling of their stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Totani2008}, \citet{Maoz2011}, \citet{Maoz2012}, \citet{Graur2013}, and \citet{Graur2015} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$.

@@

The majority of observational evidence based on studies of SN Ia populations thus points towards a DD scenario for most, if not all, SNe Ia. However, finding self-consistent progenitor and explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs (linked strongly with the SD scenario but also with many DD scenarios) find difficulty in reproducing the light curves of "normal" SNe Ia as well as "peculiar" objects (\citealt{Ropke2007,Sim2013,Blondin2017}; see \citealt{Maoz2014,Jha2019} for overviews). In recent years, attention directed towards explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations (primarily related to a DD scenario) has led to promising results \citep[e.g.][]{Shen2017,Shen2018,Townsley2019,Gronow2020,Shen2021} although they still struggle to match observations at late times in the light curve evolution \citep{Shen2021,Gronow2021}. An additional factor in support of the sub-$M_{\mathrm{Ch}}$ model is that the SN luminosity is related to the mass of the primary WD, which itself is likely to be related to its age (although this relation is probably complicated by other factors such as accretion rate, metallicity, and the composition of the companion), thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Rose2019,Nicolas2020}. Other proposed scenarios include hybrid models in which standard CO WDs merge with hybrid helium-CO WDs \citep{Zenati2019,Perets2019}. With many models showing promising similarities to observations but each subject to its own drawbacks, it is becoming accepted that more than one progenitor scenario may contribute significantly to the overall population of "normal" SNe Ia; detailed observations are thus required in order to place constraints on the relative fractions of each possible progenitor channel.

In this work, we combine parts of the traditional methods and derive a new measurement of the SN Ia DTD. Instead of measuring the volumetric rate of SNe Ia, we measure the rate per galaxy as a function of stellar mass, as per \citet{Sullivan2006,Smith2012,Brown2019}. We use the stellar mass assembly model of \citetalias{Childress2014} to predict the stellar age distribution of galaxies for a given stellar mass at the mean redshift of our SN sample, and then forward model the DTD by convolving it with the stellar age distribution and comparing the predicted rates to the observed rates.

@@

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we use Bayesian inference. To sample the posterior we use the enhanced no-U-turn Sampler (NUTS) algorithm \citep{Betancourt2017}, which is a variant of Hamiltonian Monte-Carlo (HMC), implemented in the Stan programming language \citep{Carpenter2017}. We describe the fitting procedure in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.63\pm0.02$, which is consistent with the value found in weakly star-forming galaxies in the Supernova Legacy Survey by \citet{Sullivan2006}, but also in passive galaxies in SDSS by \citet{Smith2012}, and in all galaxies in the Lick Observatory Supernova Search (LOSS) by \citet{Li2011a}. Our measured slope is also consistent with that measured in the low-redshift All-Sky Automated Survey for Supernova (ASAS-SN) Bright Supernova Catalogues \citep{Brown2019}, which extends to galaxies with stellar mass as low as $10^7$~M$_{\odot}$.

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit simultaneously by a single two-component DTD, with fixed slope $\beta = -1$ and a prompt fraction that is 1 at short delay times, 0 at large delay times, and a linear function of time between $t_1 =0.65$~Gyr (where tardy SNe switch on) and $t_2=0.85$~Gyr (where prompt SNe turn off).%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citetalias{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass for SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass, sSFR (and by inference, stellar age) as been seen by \citet{Rigault2013,Graur2017,Rigault2018,Rose2019,Hakobyan2020,Nicolas2020,Rose2021}, as well as morphology \citep{Hakobyan2020}. The corresponding DTD fit results in a steep power-law $\beta =-1.70^{+0.19}_{-0.10}$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. The difference in $\beta$ is significant at a level of $3.8\sigma$. This result provides an intriguing confirmation of that observed in 101 SDSS SNe Ia by \citet{Brandt2010} who found that low-$x_1$ SNe displayed much longer delay times than those with high $x_1$.

From the differing DTDs that describe SNe with low and high stretch values we infer that there are either multiple channels through which SN Ia explode which occur on differing timescales (as discussed in \citet{Hakobyan2020}), or that the explosion mechanism evolves with progenitor age. Two scenarios, that are not necessarily mutually exclusive, that could lead to different DTD decay rates are as follows. 
The first such scenario is to assume that all SNe Ia come from the same initial population of stars, and thus that the progenitors of both low and high stretch SNe form at the same time. It is then necessary to invoke models in which the WD progenitors of low-stretch SNe evolve on longer timescales than those of high-stretch SNe, for example due to a different initial binary separation distribution or accretion rate. In this case, the DTD slopes would indeed take on the different values measured here.

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are mildly consistent with one another, with $t_1 =0.69^{+0.07}_{-0.13}$~Gyr and $t_2 = 0.81^{+0.13}_{-0.10}$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are reminiscent of the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while low-stretch SNe occurred with delay times $\gtrsim 2.4$~Gyr. The transition found in our analysis appears to occur at much shorter delay times than the several Gyrs found by \citet{Rose2019}. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021,Nicolas2020}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but suggest that such a scenario is unlikely.

@@

One possibility is that sub-luminous SNe are more numerous in predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010,Barkhudaryan2019} at delay times as long as $>6$~Gyr \citep{Panther2019}, and are typically sub-luminous with low stretch \citep{Gonzalez-Gaitan2014}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, their presence in high-mass hosts could explain the apparent lack of normal SNe Ia in the sample. We check this hypothesis briefly by re-examining the objects left out of our sample due to \texttt{SALT2} cuts. We find that 75\% of objects left out lie in hosts with $M_*>10^{10}$ M$_{\odot}$, while this percentage is 65\% for the objects that pass the cuts. However, of the objects in high-mass galaxies that fail the cuts, only 11 (10\%) fail due to their $x_1$ lying below $-3$, and a total of 20 have $x_1 <-2$ which is the regime populated by SN 1991bg-like objects which is well short of the factor of two by which the DTD model over-predicts the data in the highest mass bin. Another issue is that SN 1991bg-like SNe are generally fainter than normal SNe Ia and the massive galaxies in which they explode are typically bright, leading to a lower detection efficiency than that modelled in Section \ref{subsec:incompleteness_SNe}. A thorough investigation into the presence of SN Ia sub-types in the DES-SN data set is deferred to a future investigation.
